---
title: "Troubleshooting"
vignette: >
  %\VignetteIndexEntry{Troubleshooting}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
    eval: false
---

When deploying applications with `buoyant`, you may occasionally encounter issues. This guide helps you diagnose and fix the most common problems, from connection failures to application crashes.

## How to Use This Guide

This troubleshooting guide is organized by problem type. To find help:

1. **Identify your problem**: Look at the section headings to find what matches your issue
2. **Check the symptoms**: Each section describes what you'll see when the problem occurs
3. **Follow the solutions**: Try the diagnostic commands and fixes provided
4. **Use the checklist**: If you're not sure where to start, see the [Debugging Checklist](#debugging-checklist)

## Quick Diagnostic Tips

Before diving into specific problems:

- **Check the logs first**: Most issues show up in the logs
  ```{r}
  droplet_ssh(droplet, "journalctl -u server-myapp -n 50")
  ```

- **Verify the service is running**:
  ```{r}
  droplet_ssh(droplet, "systemctl status server-myapp")
  ```

- **Test locally before deploying**: Use `validate_server_yml()` to catch configuration errors early

## Common Issues at a Glance

| Problem | Likely Cause | Quick Fix |
|---------|-------------|-----------|
| Can't access application | Service not running | Restart service or check logs |
| "Connection refused" | Port/firewall issue | Check port with `netstat -tlnp` |
| Application crashes | Missing dependencies | Install packages with `do_install_server_deps()` |
| Nginx errors | Configuration syntax | Run `nginx -t` to test config |
| SSL certificate fails | DNS not propagated | Wait 5-60 minutes for DNS |

## Connection Issues

### Droplet Connection Refused

If you can't connect to your droplet:

```{r}
# Refresh droplet information
droplet <- droplet(droplet$id)

# Check IP is accessible
ip <- do_ip(droplet)
system(paste("ping -c 3", ip))
```

**Common causes:**
- Droplet is still provisioning (wait a few minutes)
- Firewall blocking connections
- SSH keys not configured properly

### Can't Access Application

If your application URL isn't responding:

```{r}
# Check if the service is running
droplet_ssh(droplet, "systemctl status server-myapp")

# Check if nginx is running
droplet_ssh(droplet, "systemctl status nginx")
```

## Application Not Starting

### Check Service Logs

The most useful troubleshooting tool is the service logs:

```{r}
# View recent logs
droplet_ssh(droplet, "journalctl -u server-myapp -n 50")

# Follow logs in real-time
droplet_ssh(droplet, "journalctl -u server-myapp -f")
```

### Common Issues

**Missing Dependencies**

If you see errors about missing packages:

```{r}
# Install R packages on the server
do_install_server_deps(droplet, packages = c("dplyr", "ggplot2"))

# Or install system libraries
droplet_ssh(droplet, "sudo apt-get install -y libcurl4-openssl-dev")
```

**Port Conflicts**

If the port is already in use:

```{r}
# Check what's using the port
droplet_ssh(droplet, "netstat -tlnp | grep 8000")

# Use a different port for your application
do_deploy_server(droplet, "myapp", "path/to/app", port = 8001)
```

**Invalid `_server.yml`**

Validate your configuration locally before deploying:

```{r}
# Check your _server.yml syntax
validate_server_yml("my-app", check_engine = TRUE, verbose = TRUE)

# Read and inspect configuration
config <- read_server_yml("my-app")
str(config)
```

### Check Service Exists

Verify the systemd service was created:

```{r}
# List all server services
droplet_ssh(droplet, "systemctl list-units | grep server-")

# Check specific service status
droplet_ssh(droplet, "systemctl status server-myapp")
```

### Restart Service

Sometimes a restart helps:

```{r}
# Restart the service
droplet_ssh(droplet, "sudo systemctl restart server-myapp")

# Check status after restart
droplet_ssh(droplet, "systemctl status server-myapp")
```

## Nginx Issues

### Test Nginx Configuration

```{r}
# Test nginx configuration for syntax errors
droplet_ssh(droplet, "nginx -t")

# If errors, view the configuration
droplet_ssh(droplet, "cat /etc/nginx/sites-available/server-myapp")
```

### Check Nginx Logs

```{r}
# Check nginx error log
droplet_ssh(droplet, "tail -f /var/log/nginx/error.log")

# Check nginx access log
droplet_ssh(droplet, "tail -f /var/log/nginx/access.log")
```

### Restart Nginx

```{r}
# Restart nginx
droplet_ssh(droplet, "sudo systemctl restart nginx")

# Check status
droplet_ssh(droplet, "systemctl status nginx")
```

## Validation Errors

### Local Validation

Before deploying, validate your application:

```{r}
# Basic validation
validate_server_yml("my-app")

# Validate and check that engine is installed locally
validate_server_yml("my-app", check_engine = TRUE)

# Verbose output for debugging
validate_server_yml("my-app", check_engine = TRUE, verbose = TRUE)
```

### Common Validation Issues

**Missing Engine Field**

Your `_server.yml` must have an `engine` field:

```yaml
engine: plumber2
```

**Invalid YAML Syntax**

Check for proper indentation and formatting:

```yaml
# Correct
engine: plumber2
options:
  host: 0.0.0.0
  port: 8000

# Incorrect (bad indentation)
engine: plumber2
options:
host: 0.0.0.0
port: 8000
```

## Deployment Failures

### Upload Failures

If files fail to upload:

```{r}
# Check SSH connection
droplet_ssh(droplet, "echo 'Connection works'")

# Check available disk space
droplet_ssh(droplet, "df -h")

# Try deploying again with overwrite
do_deploy_server(droplet, "myapp", "path/to/app",
                 port = 8000, overwrite = TRUE)
```

### Package Installation Failures

If the engine package fails to install:

```{r}
# Check R is installed
droplet_ssh(droplet, "R --version")

# Try installing the package manually
droplet_ssh(droplet, "sudo R -e \"install.packages('plumber2')\"")

# Check for missing system dependencies
droplet_ssh(droplet, "sudo apt-get update && sudo apt-get install -y r-base-dev")
```

## HTTPS/SSL Issues

### Certificate Not Issued

If Let's Encrypt certificate fails:

```{r}
# Check certbot logs
droplet_ssh(droplet, "cat /var/log/letsencrypt/letsencrypt.log")

# Verify DNS is configured correctly
# (must wait for DNS propagation, usually 5-60 minutes)
system(paste("nslookup", "api.yourdomain.com"))
```

**Common causes:**
- DNS not propagated yet (wait and try again)
- Domain doesn't point to droplet IP
- Rate limit hit (Let's Encrypt has rate limits)

### Certificate Renewal Issues

```{r}
# Test certificate renewal
droplet_ssh(droplet, "sudo certbot renew --dry-run")

# Check renewal timer status
droplet_ssh(droplet, "systemctl status certbot.timer")
```

## Performance Issues

### High Memory Usage

```{r}
# Check memory usage
droplet_ssh(droplet, "free -h")

# Check which processes are using memory
droplet_ssh(droplet, "ps aux --sort=-%mem | head -n 10")

# Consider upgrading to a larger droplet
droplet <- do_provision(size = "s-2vcpu-4gb")
```

### Slow Response Times

```{r}
# Check CPU usage
droplet_ssh(droplet, "top -bn1 | head -20")

# Check if application is under heavy load
droplet_ssh(droplet, "journalctl -u server-myapp -n 100 | grep -i error")
```

## Debugging Checklist

When something goes wrong, work through this checklist:

1. **Can you connect to the droplet?**
   ```{r}
   droplet_ssh(droplet, "echo 'Connected'")
   ```

2. **Is the service running?**
   ```{r}
   droplet_ssh(droplet, "systemctl status server-myapp")
   ```

3. **What do the logs say?**
   ```{r}
   droplet_ssh(droplet, "journalctl -u server-myapp -n 50")
   ```

4. **Is the port listening?**
   ```{r}
   droplet_ssh(droplet, "netstat -tlnp | grep 8000")
   ```

5. **Is nginx working?**
   ```{r}
   droplet_ssh(droplet, "nginx -t && systemctl status nginx")
   ```

6. **Is the `_server.yml` valid?**
   ```{r}
   validate_server_yml("my-app", check_engine = TRUE, verbose = TRUE)
   ```

## Getting More Help

If you're still stuck:

1. **Check the logs** - They usually contain the answer
2. **Search existing issues** - <https://github.com/posit-dev/buoyant/issues>
3. **File a new issue** with:
   - Your `_server.yml` file
   - Service logs (`journalctl -u server-myapp -n 100`)
   - Steps to reproduce the problem
4. **Review the documentation**:
   - [Getting Started](buoyant.html)
   - [Advanced Usage](advanced-usage.html)
   - [Function Reference](https://posit-dev.github.io/buoyant/reference/)

## Related Resources

- [DigitalOcean troubleshooting](https://docs.digitalocean.com/products/droplets/troubleshooting/)
- [analogsea documentation](https://github.com/sckott/analogsea)
- [systemd documentation](https://www.freedesktop.org/software/systemd/man/)
- [nginx documentation](https://nginx.org/en/docs/)
